Planck and the Extra-terrestrials\\ or what the new SI means for astronomy

### 2020-04-08 spg/main.tex
first draft for SPG

\begin{abstract}
New freedom.
\end{abstract}

@@

Planck was by no means the only person to try

Not the first \cite{Tomilin1998} but most prescient.  SI
\cite{Jeckelmann_2018}.  Leaves out $G$.  Also candela, sievert.

Astros use Planckian units occasionally\cite{magicenv} (with $\hbar$)
but generally not.\cite{dodd2011}  Why?

Solar mass in kg would be bad.  SI in astro has been a non-starter.
But spacecraft dynamics uses $GM$ in SI.

New SI new freedoms.  eV not a new unit.  Density as
$\!\unit{eV}\unit{m}^{-3}$ is fine.  Anarchy?  No, because conversions
only through physical constants.

For astronomy, the good news is that the SI no longer requires you to
express the solar mass in kg --- but we'll get to that.  We remark
first that distances in light-seconds are compliant with the new SI.
Like the electronvolt, a light-second is not itself a unit, it is an
abbreviated way of referring to the equivalent light travel time.  On
the other hand, which astronomer would wish to lose the cultural
legacies of the astronomical unit of length (au), and the parsec?
Fortunately, metrology and cultural legacies can be reconciled.  To
see how, we only to see how modern German has kept the old unit of a
Pfund, but has rounded its meaning to exactly $500\unit{g}$.  The au
and parsec have serendipitously round values: $1\unit{au}\simeq
c\times500\unit{s}$ to better than 1\%, while a parsec is just 3\%
over $c\times10^8\unit{s}$.

@@

Now we come to the kilogram, or avoiding the kilogram when necessary.
Spacecraft dynamics has long used the so-called solar mass parameter
$GM_\odot$ in $\!\unit{m}^3\unit{s}^{-2}$, which is known to eight
significant digits in Newtonian gravity, two more with general
relativity.  In pulsar timing, $GM_\odot/c^3$ in seconds is usual.  In
the new SI, by analogy with electronvolts, we can meaningfully measure
mass as $GM/c^3$ in ``gravity-seconds''.  The solar mass is about
5~micro-(gravity)-seconds.  This does not mean that kilograms are to
be avoided in astrophysics.  Rather, gravity-seconds can be useful
where kilograms would needlessly propagate the uncertainty in $G$,
which is typically the case for orbital processes.  Together with
distances in light-seconds and dimensionless velocities,
gravity-seconds allow for some elegant simplifications in the
description of all kinds of orbital processes, classical and
relativistic.\cite{2020PASP..132b1001S} Some constructions arise, that
seem {\em very} strange at first: in particular density in
gravity-seconds per cubic light-seconds, or power in gravity-seconds
per second.  But these are not really unphysical at all.  In
gravitational phenomena, a density $\rho$ is always associated with a
time scale $(G\rho)^{-1/2}$, so density as frequency squared does make
sense.  As for dimensionless power in gravity-seconds per second, this
is just power in units of the well-known Planck power $c^5/G$, which
is the luminosity scale in gravitational waves for merging black
holes, irrespective of their mass.

>>>>

%\begin{abstract}
%\end{abstract}

@@

Planck was not the only one to propose a system of units based on
fundamental constants,\cite{Tomilin1998} but he was the most
prescient.  In Planck's system there is no prototype mass, length,
time, or temperature.  Instead there are dimensional coefficients that
appear in multiple equations of physics, and units are defined
implicitly.  In the 21st century the International System of units
concluded implicit definition of units is not a bug, it's a feature.
In 2018 the SI decided that only the second shall be defined by a
particular physical process (a spectral line in Cs).  All other units
are defined implicitly, through defined values of physical constants,
including (but of course) Planck's constant.  Any applicable equation
of physics can be used for the realization of
units.\cite{Jeckelmann_2018}

The reforms in the SI were not especially designed to communicate with
extraterrestrials.  The decision to adopt implicit definitions was
pragmatic, in that explicit definitions were sometimes overtaken by
technology.  In particular, the kilogram is best realized as a derived
electrical unit, with mass measured by way of the quantum Hall effect
and the Josephson effect.  That would be sternly frowned upon by the
old SI, but the new SI is cheerfully agnostic.  The new SI also
removes conflicts with some common scientific usages.  For example,
the electronvolt is not an SI unit.  However, if we understand mass in
electronvolts as a shorthand for the value of mass times $c^2/e$ in
volts, there is no conflict with the new SI.

Planckian units can be useful useful for theoretical work, provided
one changes to standard units for comparing with observable
quantities.\cite{magicenv} Meanwhile, the community of scientists
engaged with extraterrestrial things have steadfastly resisted
adopting SI units.  Even the proponents of SI units among astronomers
\cite{dodd2011} have not advocated a complete change to the SI.  Why
is that?  The answer may lie to some extent in sociological factors,
but there has been a scientific reason.  The key is the constant that
Planck included but the new SI does not: $G$.  The gravitational
constant is uniquely weak and uniquely hard to measure, and even now
is known to only four significant digits.\cite{2018Natur.560..562S}
The so-called solar mass parameter $GM_\odot$ is known to eight
significant digits in Newtonian gravity, two more with general
relativity.  But if you split that product into $G$ and $M_\odot$, in
order to write the mass of the Sun in kilograms, your value of the
gravitational field will be only good to four significant digits.  You
cannot navigate a spacecraft on four significant digits.  End of story
for SI in astronomy.  Or so it used to be.

But post 2018 the SI no longer requires you to express the solar mass
in kg --- but we'll get to that.  We remark first that distances in
light-seconds are compliant with the new SI.  Like the electronvolt, a
light-second is not itself a unit, it is an abbreviated way of
referring to the equivalent light travel time.  On the other hand,
which astronomer would wish to lose the cultural legacies of the
astronomical unit of length (au), and the parsec?  Fortunately,
metrology and cultural legacies can be reconciled.  To see how, we
only to see how modern German has kept the old unit of a Pfund, but
has rounded its meaning to exactly $500\unit{g}$.  The au and parsec
have serendipitously round values: $1\unit{au}\simeq
c\times500\unit{s}$ to better than 1\%, while a parsec is just 3\%
over $c\times10^8\unit{s}$.

@@

Now we come to the kilogram, or avoiding the kilogram when necessary.
Spacecraft dynamics has long used $GM_\odot$ in
$\!\unit{m}^3\unit{s}^{-2}$.  In pulsar timing, $GM_\odot/c^3$ in
seconds is usual.  In the new SI, by analogy with electronvolts, we
can meaningfully measure mass as $GM/c^3$ in ``gravity-seconds''.  The
solar mass is about 5~micro-(gravity)-seconds.  This does not mean
that kilograms are to be avoided in astrophysics.  Rather,
gravity-seconds can be useful where kilograms would needlessly
propagate the uncertainty in $G$, which is typically the case for
orbital processes.  Together with distances in light-seconds and
dimensionless velocities, gravity-seconds allow for some elegant
simplifications in the description of all kinds of orbital processes,
classical and relativistic.\cite{2020PASP..132b1001S} Some
constructions arise, that seem {\em very} strange at first: in
particular density in gravity-seconds per cubic light-seconds, or
power in gravity-seconds per second.  But these are not really
unphysical at all.  In gravitational phenomena, a density $\rho$ is
always associated with a time scale $(G\rho)^{-1/2}$, so density as
frequency squared does make sense.  As for dimensionless power in
gravity-seconds per second, this is just power in units of the
well-known Planck power $c^5/G$, which is the luminosity scale in
gravitational waves for merging black holes, irrespective of their
mass.
